### 2023-03-02 calculus.tex
Add section on marking holes

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the marked lambda calculus through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
hole}, which are used to model syntactically incomplete programs---but are not critical within the
calculus.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended gradually typed lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

@@

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally several kinds of
\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
denote a different error message that might be shown in an editor or emitted by a compiler.
%
% \begin{itemize}
  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    % $x$, as in \cref{fig:calculus-examples-unbound}.
%
  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    % expression does not match an expected type.
%
  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    % that the branches of a conditional have mismatched types.
%
  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
    % the type ascription $\TMV$ is mismatches with an expected type.
%
  % \item $\ECSynNonMatchedArrow{\ECMV}$
% \end{itemize}
%
Throughout the remainder of this section, as we formulate marking for the gradually typed lambda
calculus, we will motivate and give precise semantics for these error marks. Note that this is a
non-exhaustive list; depending on the syntactic forms and typing features found in the language,
other kinds of marks may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of
marks that may be required in an language with explicit polymorphism.

@@

To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which we
discuss next, we need only define a synthesis rule for unmarked numbers, in which numbers synthesize
the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

@@

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAHole}

@@

@@

% TODO non-gradual semantics?

% Note that while the semantics that follow concern a gradually typed language, a non-gradual system
% may be recovered straightforwardly by removing the unknown type and replacing type consistency with
% equality.

@@

  \judgbox{\ensuremath{\matchedProd{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TProd{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
the Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we
then consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the marked lambda calculus through its application to the GTLC extended
with numbers and booleans. Consider the syntax for such a language, given as $\EMName$ in
\cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond
to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. Finally, $\EEHole$ denotes an
\emph{expression hole}, which are used to model syntactically incomplete programs. Note, however,
that these are not critical to the calculus.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended GTLC.}
  \label{fig:calculus-syntax}
\end{figure}

@@

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally several kinds of
\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
denote a different error message that might be shown in an editor or emitted by a compiler.
%
% \begin{itemize}
  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    % $x$, as in \cref{fig:calculus-examples-unbound}.
%
  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    % expression does not match an expected type.
%
  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    % that the branches of a conditional have mismatched types.
%
  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
    % the type ascription $\TMV$ is mismatches with an expected type.
%
  % \item $\ECSynNonMatchedArrow{\ECMV}$
% \end{itemize}
%
Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
give precise semantics for these error marks. Note that this is a non-exhaustive list; depending on
the syntactic forms and typing features found in the language, other kinds of marks may be
necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be required
in an language with explicit polymorphism.

@@

To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which is
discussed in more detail next, we need only define a synthesis rule for unmarked numbers, in which
number literals synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

@@

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAUnknown}

@@

\subsubsection{Holes}
For completeness, we finish the development of the marked lambda calculus on the GTLC with the
marking of holes. Like number literals and booleans, these are governed by straightforward rules.
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
  }{USHole}

  \judgment{ }{
    \ctxSynFixedInto{\ctx}{\EEHole}{\ECEHole}{\TUnknown}
  }{MKSHole}

  \judgment{ }{
    \ctxSynTypeM{\ctx}{\ECEHole}{\TUnknown}
  }{MSHole}
\end{mathpar}

@@

@@

  \judgbox{\ensuremath{\matchedProd{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TProd{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPUnknown}
